Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Guideline for International \emph{and Domestic} Students

### 2024-02-10 demystify.tex
updates

Most PhD programs will require at least \textbf{two LORs}. Having a letter from an internationally recognized researcher can greatly strengthen your application. However, obtaining such letters
can be challenging for international students, who might not have much interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, postdocs who mentored you) that know you well enough to talk about \emph{your research experience and capabilities}. Many students get letters from supervisors from companies where they did internships or are working at. It is OK as long as it is a research-based personalized letter (once again, we are talking about PhD applications, not MS).

@@

As mentioned, if you're admitted to a \emph{good} CS PhD program, you should not have to worry about funding!  
In the US, the common types of funding for PhD are \emph{graduate teaching assistant} (GTA or TA), \emph{graduate research assistant} (GRA or RA), and \emph{Fellowship}.
RA is paid by a prof. for you to do their research. TA is paid by the department for you to help with teaching. Finally, fellowship is an independent funding that can come from a school, a company, or an organization. \autoref{tab:funding} summarizes the differences.
Note that funding is typically more available for PhD students than 
Masters. %Undergraduates typically have no funding and also have to pay international tuition, which is very expensive!.

@@

\section{Graduate Assistantship (TA/RA)}
The most common type of funding is \textbf{graduate assistantship}, which is either TA or RA. Both TA and RA come with tuition waiving (you don't have to pay tuition), health insurance (this takes care of your insurance, which is a must have in the US), and most importantly, your stipend (i.e., your salary). Some universities also pay insurance for spouse/children (or give very good discount).

Several things about stipends.  First, the amount of stipend depends on the university, which in turn depends on various factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living cost). Second, a school year is typically 9-month in the US, so stipend is for 9 months (so divide by 9 for each month). Third, like most source of income in the US, you will have to pay tax on your stipend.  Fourth, CS department typically has higher stipend comparing to other study fields.  Finally, private universities might pay more for stipend (but they could have higher "activity" or some other hidden fee, or you may be required to pay fees for each credit hour).

@@

Fellowship is another type of funding that students can apply for (e.g., from school, industries, government). Fellowships are typically competitive and generous, and gives pretty much all benefits tuition/insurance that a TA/RA has.  Moreover, they often give higher stipend (including summer) and open doors for job opportunities (e.g., internship).

In general, fellowship is prestigious, and you will stand out if you get one.  Every PhD student has pubs, but only superstars have NSF grad or Microsoft fellowship. In fact, these are so prestigious that even if you didn't get it but make it to the final round, school will still mention you on their website and you still should put it on your CV.

\textbf{How to get Fellowship?} You need to apply for them.  The US government has many fellowships, though they would likely require US citizenship or residency.  However, tech companies including Google, Microsoft, Facebook have fellowships that international students can apply for.

Prestigious fellowships typically require a clear and good research plan, so it is a good idea to wait until at least your second year to have research experience and even publication before applying. Remember, you're competing with the top PhD students at top universities worldwide.

\begin{commentbox}[Vu:]
  PhD applicants at GMU are automatically eligible for a \emph{Presidential Fellowship}.  It is at least as good as GTA but the most important thing is that as a fellowship it is truly free money (i.e., you are not depending on any prof. or TA).  PhD adcom members nominate applicants for this fellowship and the committee will vote and give the fellowship to the top 2.
\end{commentbox}

>>>>

Most PhD programs will require at least \textbf{two LORs}. Having a letter from an internationally recognized researcher will greatly strengthen your application. In fact, a strong letter from a well-known researcher can outweigh the lack of publications.

However, obtaining such letters
can be challenging for international students, who might not have much interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, postdocs who mentored you) that know you well enough to talk about \emph{your research experience and capabilities}. Many students get letters from supervisors from companies where they did internships or are working at. It is OK as long as it is a research-based personalized letter (once again, we are talking about PhD applications, not MS).

@@

As mentioned, if you're admitted to a \emph{good} CS PhD program, you should not have to worry about funding!  
In the US, the common types of funding for PhD are \emph{graduate teaching assistant} (GTA or TA), \emph{graduate research assistant} (GRA or RA), and \emph{Fellowship}.
RA is paid by a prof. for you to do their research. TA is paid by the department for you to help with teaching. Finally, fellowship is an independent funding that can come from a school, a company, or an organization. \autoref{tab:funding} summarizes the differences.
Note that funding is typically more available for PhD students than 
Masters.

@@

\section{Graduate Assistantship (TA/RA)}
The most common type of funding is \textbf{graduate assistantship}, which is either TA or RA. Both TA and RA come with tuition waiving (you don't have to pay tuition), health insurance (this takes care of your insurance, which is a must have in the US), and most importantly, your stipend (i.e., your salary). Some universities also give significant discount or pay insurance for spouse/children.

Several things about stipends.  First, the amount of stipend \emph{varies} and depends on various factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living cost). Second, an academic year (AY)  year is typically \emph{9-month} in the US, so stipend is for 9 months (so divide by 9 for each month). Third, you might get paid over the summer through funding from your professor or fellowship (typicaly no TA over the summer). Fourth, like most source of income in the US, you will have to \emph{pay tax} on your stipend.  Finally, private universities might pay more for stipend, but they could have higher "activity" or some other hidden fee, or you may be required to pay fees for each credit hour.

@@

Fellowship is another type of funding that students can get (e.g., from the university, industries, government).
Fellowships are typically competitive and generous, and gives pretty much all benefits tuition/insurance that a TA/RA has.  Moreover, they often give higher stipend (including summer) and open doors for job opportunities (e.g., internship).

\textbf{How to get Fellowship?} You need to apply for them.  Many schools provide such fellowships to attract admitted students.
External fellowships including from the US government and tech companies (e.g., Google, Microsoft, Facebook).  Fellowships from US government would likely require US citizenship or residency. % However, tech companies including Google, Microsoft, Facebook have fellowships that international students can apply for.

Prestigious external fellowships typically require a clear and good research plan, so it is a good idea to wait until at least your second year to have research experience and even publication before applying. Remember, you're competing with the top PhD students at top universities worldwide.

\begin{commentbox}[Vu:]
  PhD applicants at GMU are automatically eligible for a \emph{Presidential Fellowship}.  It is at least as good as GTA but the most important thing is that as a fellowship it is truly free money (i.e., you are not depending on any prof. or TA).  PhD adcom members nominate applicants for this fellowship and the whole committee will decide.
\end{commentbox}

In general, fellowship, especially competitive ones from the government or company, is prestigious, and you will stand out if you get one.  Every PhD student has pubs, but only superstars have NSF grad or Microsoft fellowship. In fact, these are so prestigious that even if you didn't get it but make it to the final round or even ``honorable mentioning'', school will still mention you on their website and you still should put it on your CV.
